---
title: "WGS Unknown NTM_22-11 Investigation"
output:
     pdf_document: default
     html_document: default
header-includes:
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{xcolor}
  - \usepackage{colortbl}
---

## Preface


We have recieved unknown samples of Non-tuberculous mycobacteria (NTM) collected in Hawai'i, a region with the highest 
reported incidence of NTM infection in the United States. NTMs are a diverse group of over 200 speicies of bacteria found in
soil,freshwater,facuets and showerheads, where they can multiply. Exposure to these types of environments, particularly in Hawaii's humid environment, 
promotes NTM survival and transimission, posing as a huge risk for the public. The National Institute of Health stated that Hawaii has two-fold
greater than the national average risk of infection, causing issues in the lungs. Despite the dangers of NTMS, the diversity of NTMs
in Hawai'i is still incompletely discovered. A clinical laboratory in Hawaii collected NTM_22-11 as a potentially novel NTM. 
In order to confirm the identity of the samples, we have performed whole genome sequencing (WGS) on the samples using the Oxford Nanopore MinION sequencing platform.
Samples were prepared using the rapid barcoding kit (SQK-RBK114.24) and sequenced on a MinION flow cell.
Raw sequencing data were basecalled and demutiplexed with Doroado v0.8.2, where reads were filtered using Chopper.
Draft genomes were assembled using Flye v2.8.3 and polished using Medaka v1.0.3. Genomic characterization of NTM_22-11 will help
us discover potentially novel speicies that affect the public health.

```{r warning=FALSE, message=FALSE}
library(stringr)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(ape)
```

## Completeness Check
Here, we want to measure the genome completeness. Using Docker and CheckM, it allows us to estimate how much of the 
expected bacterial genome is present by searching genes that should exsist in all bacteria. Where the clincial labatory in Hawaii
collected samples, there is a possibility of other environmental bacteria present in the data. CheckM flags contamination so that
genome size isn't inflated. Overall Checkm is used to make sure NTM_22-11 is accurately represented as a single bacteria.


```{sh eval=FALSE}
# Running checkM
# Place the docker command for checkM here
docker pull nanozoo/checkm
docker run --name=checkm \
--rm \
--volume=$PWD:/home \
--workdir=/home \
-it nanozoo/checkm:latest

checkm taxonomy_wf domain Bacteria -t 2 -x fasta ./ ./checkm_out

checkm qa -o 2 --tab_table -t 2 \
-f checkm_out/summary.tsv checkm_out/Bacteria.ms checkm_out

```


```{r}
# Read in the checkM summary
checkm <- read.csv("checkm_out/summary.tsv", sep = "\t")

# Create tables for checkM completion and contamination
# Make sure to include the following fields: Bin.Id, Marker.lineage, Completeness, Contamination
# Make sure to rename the columns to ID, Lineage, Completeness, Contamination
checkm_table <- data.frame(checkm)
checkm_table_renamed <- checkm_table %>%
  select(ID = Bin.Id, Lineage = Marker.lineage, Completeness = Completeness, Contamination = Contamination)

# Create table for checkM additional summary stats
# Make sure to include the following fields: Bin.Id, Genome.size..bp., GC, X..contigs, N50..contigs.
# Make sure to rename the columns to ID, Genome Size (bp), GC, Contig #, N50
checkm_stats <- checkm_table %>%
select(Bin.Id, Genome.size..bp., GC, X..contigs, N50..contigs.) %>%
rename(ID = Bin.Id, Genome_Size_bp = Genome.size..bp., Contigs = X..contigs, N50 = N50..contigs.)
```

```{r}
# Print tables
# checkm_table_renamed %>%
#   kable("latex", booktabs = TRUE, caption = "CheckM Overview",
#         col.names = "Overview stats") %>%
#   kable_styling(latex_options = c("striped", "hold_position"))

kable(checkm_table_renamed)

# checkm_stats %>%
#   kable("latex", booktabs = TRUE, caption = "CheckM Summary Stats",
#         col.names = "Summary stats") %>%
#   kable_styling(latex_options = c("striped", "hold_position"))

kable(checkm_stats)
```

## Prokka Annotation
The next step is to turn raw DNA sequences into something that is readable. The genome contain the string of "A","T","C","G". Prokka
allows us to look at the tRNA, rRNA and other elements within the genome to conduct an analysis. Prokka also makes sure our NTM is novel since the gene content
allows us to distinguish between different NTMs. 

```{sh eval=FALSE}
# Running Prokka
# Place the docker command for Prokka here
docker run --name=prokka \
--rm \
--volume=$PWD:/data \
-it staphb/prokka:latest

prokka --outdir prokka_out --prefix prokka_annotation \
--cpus 8 NTM_22-11.fasta
```


```{r}
# Read in the prokka annotation
prokkat <- read.csv("prokka_out/prokka_annotation.tsv",sep = "\t")
# Summarize the prokka annotation
# You will need to count the number of CDS, tRNA, rRNA, and non-hypothetical CDS
# Number of genes annotated
#prokkat$ftype[1]
CDS_count <- 0
rRNA_count <- 0
tRNA_count <- 0
non_hypoCDS_count <- 0 
for(i in 1:nrow(prokkat)){
  if(prokkat$ftype[i] == "CDS"){
    CDS_count <- CDS_count + 1
  }else if(prokkat$ftype[i] == "rRNA"){
    rRNA_count <- rRNA_count + 1
  }else if(prokkat$ftype[i] == "tRNA"){
    tRNA_count <- tRNA_count + 1
  }
}

for(i in 1:nrow(prokkat)){
if(prokkat$product[i] != "hypothetical protein"){
non_hypoCDS_count <- non_hypoCDS_count + 1
}

}
  CDS_count
# Combine the counts into a table
# Make sure to clear the rownames for clean formatting 
prokka_summary <- data.frame(
CDS = CDS_count,
tRNA = tRNA_count,
rRNA = rRNA_count,
non_hypoCDS = non_hypoCDS_count
)
```

```{r}
# Print table
kable(prokka_summary)
# prokka_summary %>%
#   kable("latex", booktabs = TRUE, caption = "Prokka Annotation Summary") %>%
#   kable_styling(latex_options = c("striped", "hold_position"))
```


```{r}
# Check to see if 16S is present & extract if present
#use ape
prokka <- read.FASTA("prokka_out/prokka_annotation.ffn")
# class(prokka) 
# str(prokka)
# names(prokka)
RNA <- prokka[grep("16S ribosomal", names(prokka))]
write.FASTA(RNA, file = "16S_unkownNTM.fasta")

```


## BLAST 16S rRNA Identification
The 16S rRNA gene is used for identifying bacterial because it is in almost all bacteria, has conserved regions that is unique to 
each bacteria. Running BLASTn with our bacteria determines whether NTM_22-11's genomic representation is close to other
bacteria speicies. When a result shows a high similarity to other bacteria but no exact matches, it tells us 
that NTM_22-11 could be novel. 

```{sh eval=FALSE}
# Running BLAST
# Place the docker command for BLAST here
Docker build -t 16sblast:local .

docker run --name=blastn \
--rm \
--volume=$PWD:/home \
--workdir=/home \
-it 16sblast:local
```

```{r}
# Read in the BLAST results
blast <- read.delim("blast_results.txt", header = FALSE)
# Make sure to remove columns 4, 5, and 9
# Make sure to parse species name from column V8 and add it as a new column
# Make sure to rename the columns to Sample ID, Identity (%), Length, E-value, Bit Score, Species
drops <- c(colnames(blast[4]),colnames(blast[5]), colnames(blast[9]))
blast <- blast[, !names(blast) %in% drops]
```

```{r}
# Create table
kable(blast)
# blast %>%
#   kable("latex", booktabs = TRUE, caption = "16S BLAST Results",
#         col.names = c("Sample ID", "Identity (%)", "Length", "E-value", "Bit Score", "Species"))
```

## Phylogenetic ID based on 16S rRNA
Next, we map the similar bacteria on a Phylogenetic tree to understand the evolution of each bacteria, and how close it is 
to NTM_22-11. THe tree will visually tell us whether NTM_22-11 groups with other bacterial speicies or formed a seperate branch 
from an exsiting bacteria. Using MUSCLE in this process will allow us to align the 16s rRNA gene to create the phylogentic tree. IQ TREE allows us
to test multiple models for the best fitted model when it comes to creating the phylogentic tree. BLAST on its own isn't the most accurate way to 
tell if NTM_22-11 is novel since many bacterial speicies sharing almost the same 16S identity. The phylogentic tree gives us an representation of
the evolution of bacterias that can be similar to NTM_22-11.
```{r}
one <- read.FASTA("16S_unkownNTM.fasta")
two <- read.FASTA("modified_representative_seqs.fasta")
three <- read.FASTA("outgroup.fasta")

out <- c(one, two, three)
write.FASTA(out, "combined_1.fasta")
```

```{sh eval=FALSE}
# Running MUSCLE and IQ-TREE
# Place the docker command for MUSCLE and IQ-TREE here
#muscle
docker run --name=muscle \
    --rm \
    -v $PWD:/data \
    --workdir=/data \
    pegi3s/muscle -in /data/combined_1.fasta -out /data/musclecombined.fasta 

#tree
docker run --name=iqtree2 \
--rm \
--volume=$PWD:/home \
--workdir=/home \
staphb/iqtree2:latest iqtree2 -s musclecombined.fasta -m TEST -B 1000 -T 8
```


```{r}
# Read in the tree and root at outgroup
tree <- read.tree("musclecombined.fasta.contree")
# Use ?root for info
tree$tip.label
root <- root(tree, outgroup = "NR_044590.1_Hoyosella_altamirensis")
# Define the name of the unknown sample
unknown_sample1 <- "AOKDPPHH_03798"
unknown_sample2 <- "AOKDPPHH_03803"
unknown_sample3 <- "AOKDPPHH_04179"
unknown_sample4 <- "AOKDPPHH_04184"

# Create a vector of colors for the tips
# Make the unknown sample a different color from the rest
tip_colors <- "black"
tip_colors[str_detect(tree$tip.label, unknown_sample1)] <- "red"
tip_colors[str_detect(tree$tip.label, unknown_sample2)] <- "red"
tip_colors[str_detect(tree$tip.label, unknown_sample3)] <- "red"
tip_colors[str_detect(tree$tip.label, unknown_sample4)] <- "red"
```

```{r fig.width = 13, fig.height = 16, fig.cap = "Phylogenetic tree of the unknown NTM sample based on the 16S rRNA gene."}
# Plot the tree
plot(tree, tip.color = tip_colors, cex = 0.7)
nodelabels(tree$node.label, cex = 0.7, frame = "none", adj = c(1, -0.5))
```


## Antimicrobial Resistance Gene Identification
NTMs are usually resistant to many antibotics, thus infections are hard to treat. Using AMRFinderPLUs allows us to 
detect known antimicrobial resistance genes in NTM_22-11. This could help reserachers understand what kinds of antibiotics 
are ineffective against NTM_22-11 so that they can create an effective treatment. 
```{sh eval=FALSE}
# Running AMRFinderPlus
# Place the docker command for AMRFinderPlus here
docker pull ncbi/amr:latest

docker run --name=amrfinder \
--rm \
--volume=$PWD:/home \
--workdir=/home \
ncbi/amr:latest amrfinder \
    --threads 8 \
    -a prokka \
    -p ./prokka_out/prokka_annotation.faa \
    -g ./prokka_out/prokka_annotation.gff --print_node \
    --plus \
    --protein_output ./amr_genes/amrfinder_pro.fasta >./amr_genes/amrfinder_pro_results.tsv

```


```{r}
# Read in the AMR results
# Extract the Protein.identifier, Gene.symbol, and Sequence.name columns into a new data frame
# Rename the columns to Protein ID, Gene Symbol, and Sequence Name
amrs <- read.csv("amr_genes/amrfinder_pro_results.tsv", sep = "\t")
amr_summary <- data.frame(amrs) %>%
select (Protein.id, Element.name, Element.symbol) %>%
rename(Protein_ID = Protein.id, Gene_Symbol = Element.symbol, Sequence_Name = Element.name)

```

```{r}
kable(amr_summary)
# amr_summary %>%
  # kable("latex", booktabs = TRUE, caption = "AMR Gene Summary",
  #       col.names = c("Protein ID", "Gene Symbol", "Sequence Name")) %>%
  # kable_styling(latex_options = c("striped", "hold_position"))

```


## Conclusions
From observing the phylogentic tree, multiple Mycobacterium speicies could be found across the branches. 
This means that across most Mycobacterium speicies, it is similar to the NTM_22-11 bacteria. 
With four unkown speicies in red, they represent multiple 16s rRna gene copies from NTM_22-11 or similar contigs 
from the same genome. We also observed that the phylogentic tree is compressed because the 16s rRNA 
gene fits very closely to other Mycobacterium species, so it is hard to diffierinate NTM_22-11 from other NTMs. Thus,
there is no final conclusion on whether NTM_22-11 is a novel speicies. More in depth analysis such as nucleotide identitfication
may be used to further investigate whether NTM_22-11 is a novel bacteria species. Despite not having a defenitive answer,
there are information on what antibiotics the potential NTM is resistant to. 

## Source
https://pmc.ncbi.nlm.nih.gov/articles/PMC5631500/

```{r eval=FALSE, echo=FALSE}
library(tinytex)
library(rmarkdown)
render("WGS_report_stub.Rmd", output_format = "pdf_document")
```